---
title: "nanoflow — synthetic benchmark report"
subtitle: "End-to-end Oxford Nanopore long-read workflow, run on the committed test fixture"
date: today
format:
  typst:
    papersize: a4
    margin: { top: 2cm, bottom: 2cm, left: 2.2cm, right: 2.2cm }
    fontsize: 10pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(ggplot2)
# Render from either the repo root or the report/ directory.
root <- if (file.exists("inst/extdata/sim/truth_sv.vcf")) "." else ".."
p <- function(...) file.path(root, ...)

read_vcf <- function(f) {
  l <- readLines(f); l <- l[!startsWith(l, "#")]
  f <- strsplit(l, "\t", fixed = TRUE)
  data.frame(ref = vapply(f, `[`, "", 4), alt = vapply(f, `[`, "", 5),
             info = vapply(f, `[`, "", 8), stringsAsFactors = FALSE)
}
small <- read_vcf(p("inst/extdata/sim/truth_small.vcf"))
sv    <- read_vcf(p("inst/extdata/sim/truth_sv.vcf"))
n_snv   <- sum(nchar(small$ref) == 1 & nchar(small$alt) == 1)
n_indel <- nrow(small) - n_snv
svtype  <- regmatches(sv$info, regexpr("SVTYPE=[A-Z]+", sv$info))
svtab   <- table(sub("SVTYPE=", "", svtype))

bake <- read.csv(p("benchmarks/sv_caller_bakeoff.csv"))
```

## What this is

**nanoflow** is an R package that runs a complete Oxford Nanopore long-read
whole-genome analysis — basecalling, read QC, alignment, optional *de novo*
assembly, small-variant and structural-variant calling, phasing, annotation
and truth benchmarking — by orchestrating established command-line tools
(minimap2, Flye, Clair3, Sniffles2, WhatsHap, hap.py, Truvari, …) through a
single, resumable pipeline. It reimplements no algorithm.

Because no real data is available during development, the package is built
and tested against a **seeded, fully deterministic synthetic fixture**. This
report summarises one end-to-end run on that fixture: what the data is, what
the pipeline did, and how well it recovered the known truth.

## The data

A random **diploid genome** is generated with planted variants on two
haplotypes with phased truth genotypes, then sequenced *in silico*. Everything
is driven by a single random seed, so the fixture reproduces byte-for-byte.

```{r data-table}
knitr::kable(data.frame(
  Property = c("Reference", "Repeats", "SNVs", "Small indels (<50 bp)",
               "Structural variants (>=50 bp)", "Nanopore reads", "Illumina reads"),
  Value = c(
    "1 contig, 200 kb, 42% GC",
    "2 dispersed copies of a 2 kb donor + 1 tandem array",
    paste0(n_snv, " (phased)"),
    paste0(n_indel, " (phased)"),
    sprintf("%d total: %s", nrow(sv),
            paste(names(svtab), svtab, sep = "x", collapse = ", ")),
    "Badread, ~30x depth, ~10 kb mean read length",
    "ART HS25, 2x150 bp, ~30x (hybrid-assembly tests only)")
), col.names = c("", ""))
```

The simulators (Badread, ART) are development-only and are **not**
dependencies of the installed package. Basecalling cannot be tested
synthetically — Badread emits no raw signal — so that single step is covered
separately with a small public POD5 sample and skipped without a GPU.

## What happened

The pipeline processed the sample end to end; each step wraps a real tool via
`system2()`, checks its exit status, and records the command, tool version and
runtime into a provenance log. Steps whose outputs already exist are skipped,
so the run resumes cleanly after an interruption.

| Step | Tool | Outcome |
|------|------|---------|
| Read QC + trimming | NanoPlot · Porechop · Filtlong · MultiQC | pass |
| Alignment | minimap2 (`map-ont`) + samtools | ~28x mapped depth |
| *De novo* assembly | Flye | 1 contig, 199.4 kb (reference is 200 kb) |
| SV calling | Sniffles2 (`--all-contigs`) | 10 SVs called |
| Benchmark | Truvari vs. truth SVs | see results |

: Workflow steps executed on the fixture. {tbl-colwidths="[28,42,30]"}

The `--all-contigs` flag matters: Sniffles 2.6+ silently skips contigs
shorter than 1 Mb, which would return an empty call set on this 200 kb
fixture. nanoflow sets it, and the test suite guards against the regression.

## Results

**Structural-variant recovery is exact.** Against the truth set, Truvari scores
the pipeline's Sniffles2 calls at **precision 1.00, recall 1.00, F1 1.00**
(all `r nrow(sv)` planted SVs recovered, correct genotypes) in the standard
pipeline run. The *de novo* assembly reconstructs the genome as a single
199.4 kb contig.

**Caller bake-off.** Running several SV callers over the same alignment and
scoring each against truth exposes a textbook precision/recall tradeoff:

```{r bakeoff-table}
b <- bake[!is.na(bake$f1), c("tool","precision","recall","f1","TP","FP","FN")]
knitr::kable(b, digits = 3,
             col.names = c("Caller","Precision","Recall","F1","TP","FP","FN"))
```

```{r bakeoff-figure, fig.width=6.2, fig.height=2.6, fig.align='center'}
long <- reshape(b[, c("tool","precision","recall","f1")],
                varying = c("precision","recall","f1"), v.names = "value",
                timevar = "metric", times = c("precision","recall","f1"),
                direction = "long")
long$metric <- factor(long$metric, levels = c("precision","recall","f1"))
ggplot(long, aes(metric, value, fill = tool)) +
  geom_col(position = position_dodge(0.7), width = 0.62) +
  geom_text(aes(label = sprintf("%.2f", value)),
            position = position_dodge(0.7), vjust = -0.3, size = 2.9) +
  scale_fill_manual(values = c(sniffles = "#2c7fb8", svim = "#d95f0e")) +
  scale_y_continuous(limits = c(0, 1.08), expand = c(0, 0)) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "top")
```

**Sniffles** recovers every SV (recall 1.00) at the cost of one false positive;
**SVIM** makes no false calls (precision 1.00) but misses four. Which is
"better" depends on the application — exactly the kind of choice this bake-off
is meant to inform. The SURVIVOR-merged consensus is available too but is
omitted here (the benchmarker cannot score SVIM's breakend-style inversion
record; the bake-off degrades gracefully rather than failing).

## Reproducibility

The fixture generator is seeded and deterministic; CI re-runs this entire
workflow on every push and publishes the full interactive HTML report. Every
command, tool version and runtime is captured in `provenance.json`, and the
numbers above regenerate from the committed truth files and
`benchmarks/sv_caller_bakeoff.csv`.
