---
title: "Real-data walkthrough: GIAB HG002 on GRCh38"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-data walkthrough: GIAB HG002 on GRCh38}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Documentation vignette: nothing here is executed at build time (the data
# is hundreds of GB and needs a GPU + cluster).
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

This vignette documents a complete real-data run on the public
Genome-in-a-Bottle **HG002** sample, from raw POD5 to benchmarked,
annotated variants. HG002 is ideal because GIAB publishes both Nanopore
raw data and high-confidence truth sets, so the same benchmarking the
pipeline applies to the synthetic fixture works on real data.

## 1. Set up a fresh server

Follow the README's "install on a new server" section: clone the
repository, create the conda environments from `inst/conda/`, install the
package, then verify:

```{r}
library(nanoflow)
check_tools()
has_gpu()   # TRUE required only for the basecalling step
```

## 2. Download the reference and databases

```{r}
refs <- download_references("/data/refs", what = c("genome", "snpeff"))
# $genome_fasta: GCA_000001405.15_GRCh38_no_alt_analysis_set.fna (+ .fai)
# $snpeff_data_dir: /data/refs/snpeff_data (GRCh38.105)
```

## 3. Get the HG002 data and truth sets

All of HG002 is openly available, with no registration or authentication.
The nanopore reads come from ONT Open Data (Open Data on AWS), and the
truth sets from NIST/GIAB. URLs below verified live 2026-10 (GIAB migrated
its FTP to `/ReferenceSamples/giab/` and the current `latest` benchmark is
**v5.0q**, which — unlike the older v4.2.1 small-variant + GRCh37-only SV
Tier1 v0.6 sets — provides both small-variant and SV benchmarks natively on
GRCh38, so the whole run stays in one coordinate system):

```bash
# ONT Open Data: R10.4.1 HG002 PromethION (POD5 + basecalled modBAM),
# anonymous S3, no credentials. Browse the release to pick a flowcell, then
# sync its POD5 directory (listing shows the exact sub-paths):
#   https://epi2me.nanoporetech.com/giab-2025.01/   (overview + layout)
aws s3 ls   --no-sign-request s3://ont-open-data/giab_2025.01/
aws s3 sync --no-sign-request s3://ont-open-data/giab_2025.01/<flowcell>/ /data/hg002/pod5/

# GIAB HG002 v5.0q truth sets on GRCh38 (small variants + SVs + regions)
G=https://ftp-trace.ncbi.nlm.nih.gov/ReferenceSamples/giab/release/AshkenazimTrio/HG002_NA24385_son/latest
wget $G/HG002_GRCh38_v5.0q_smvar.vcf.gz{,.tbi}
wget $G/HG002_GRCh38_v5.0q_smvar.benchmark.bed       # small-variant confident regions
wget $G/HG002_GRCh38_v5.0q_stvar.vcf.gz{,.tbi}
wget $G/HG002_GRCh38_v5.0q_stvar.benchmark.bed       # SV confident regions
```

(Older established alternatives, if a reviewer expects them: small variants
`NISTv4.2.1/GRCh38/` and SV `data/.../NIST_SVs_Integration_v0.6/` — but the
latter is GRCh37 only.)

Sample sheet (`/data/hg002/samples.csv`):

```
sample,reads,truth_small_vcf,truth_sv_vcf
HG002,/data/hg002/pod5/,/data/refs/HG002_GRCh38_v5.0q_smvar.vcf.gz,/data/refs/HG002_GRCh38_v5.0q_stvar.vcf.gz
```

## 4. Configure the run

`/data/hg002/config.yml`:

```yaml
reference:
  fasta: /data/refs/GCA_000001405.15_GRCh38_no_alt_analysis_set.fna
  snpeff_db: GRCh38.105
  snpeff_data_dir: /data/refs/snpeff_data
output_dir: /data/hg002/results
threads: 32
memory_gb: 128
tools:
  dorado: /opt/dorado/bin/dorado      # ONT build, GPU node
  happy: /opt/conda/envs/nanoflow-happy/bin/hap.py
steps:
  basecall: {model: "sup"}
  small_variants:
    model: r1041_e82_400bps_sup_v420   # match your flowcell/chemistry!
  sv:
    extra_args: ["--tandem-repeats", "/data/refs/human_GRCh38_no_alt.trf.bed"]
  annotate: {run: true}
  report:
    igv_regions: ["chr20:63025520-63075520"]
```

Two settings matter most on real ONT data: the **Clair3 model** and the
**Dorado model** must match the flowcell chemistry and basecalling
accuracy mode, and Sniffles benefits from the published GRCh38
**tandem-repeat annotations**.

## 5. Run

On a workstation:

```{r}
run <- run_pipeline("/data/hg002/samples.csv", "/data/hg002/config.yml")
```

On a Slurm cluster (one job per sample; basecalling needs `gpus = 1`):

```{r}
submit_slurm("/data/hg002/samples.csv", "/data/hg002/config.yml",
             partition = "gpu", gpus = 1, cpus = 32, mem_gb = 128,
             setup_lines = c("module load apptainer",
                             "source activate nanoflow"))
```

Interrupted? Just rerun — finished steps are detected and skipped.

## 6. What to expect

With ~30× R10.4.1 sup reads, published results for this tool stack are
roughly: SNV F1 ≈ 0.99+ (hap.py, GIAB confident regions), indel F1 ≈
0.8–0.9, SV F1 ≈ 0.9–0.95 (Truvari vs the v5.0q GRCh38 SV benchmark), phase block N50 in the
hundreds of kb with switch error < 1 %. The run report
(`results/report.html`) gathers hap.py/Truvari tables, WhatsHap stats,
QC and alignment metrics across samples; `results/HG002/benchmark/`
holds the raw hap.py and Truvari outputs; `provenance.json` records every
command for the methods section of a paper.

## 7. Benchmarking variations

To compare callers (Phase 2 tools), rerun with a different selection and
diff the benchmark tables — resume means only the changed steps rerun:

```yaml
steps:
  sv: {tool: svim}         # or nanovar; merge_survivor() combines call sets
  small_variants: {tool: medaka}
  phase: {tool: hapcut2}
```
