---
title: "Running the nanoflow workflow on the synthetic fixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the nanoflow workflow on the synthetic fixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# The workflow chunks only evaluate when the core external tools are
# present (e.g. inside the conda environment or the Apptainer container);
# otherwise this vignette still builds and shows the code.
core_tools <- c("minimap2", "samtools", "sniffles", "truvari",
                "bgzip", "tabix")
can_run <- all(nzchar(Sys.which(core_tools)))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = can_run)
```

nanoflow orchestrates an end-to-end Oxford Nanopore whole-genome analysis:
every step wraps an established command-line tool via `system2()`, checks
its exit status and returns a `nanoflow_step` object with output paths,
tool version, parameters and runtime. This vignette runs the pipeline on
the tiny synthetic fixture that ships with the package: a 200 kb diploid
genome with 290 planted small variants and 10 SVs, ~30× simulated Nanopore
reads, ~30× Illumina pairs, and phased truth VCFs
(see `data-raw/` in the source repository for the generator).

## The inputs

A run needs two things: a *sample sheet* and a *config*.

```{r, eval = TRUE}
library(nanoflow)

sheet_csv <- system.file("extdata/sim/sample_sheet.csv", package = "nanoflow")
read_sample_sheet(sheet_csv)[, c("sample", "input_type")]
```

The config is a YAML file (or R list) merged over `default_config()`; only
non-default values are needed. All paths — reference, databases, binaries,
output — live here, never in the code:

```{r, eval = TRUE}
cfg <- list(
  reference = list(fasta = system.file("extdata/sim/genome.fa",
                                       package = "nanoflow")),
  output_dir = file.path(tempdir(), "nanoflow_demo"),
  threads = 2,
  steps = list(
    qc = list(run = FALSE),            # skip QC here to keep the demo fast
    assembly = list(run = FALSE),
    small_variants = list(run = FALSE),# Clair3 model dir not set up here
    # Sniffles >= 2.6 skips contigs < 1 Mb unless told otherwise -- our
    # fixture contig is 200 kb, so this flag matters:
    sv = list(run = TRUE, extra_args = "--all-contigs"),
    report = list(run = FALSE)
  )
)
```

Before running anything, verify the environment:

```{r}
check_tools(tools = c("minimap2", "samtools", "sniffles", "truvari"))
```

## Running the pipeline

```{r}
run <- run_pipeline(sheet_csv, cfg)
run
```

Each sample went through alignment and SV calling; because the sample
sheet provides `truth_sv_vcf`, Truvari benchmarking ran automatically:

```{r}
run$samples$sim1$benchmark_sv$params$metrics
```

Calling `run_pipeline()` again with the same config *resumes*: steps whose
outputs exist are skipped.

```{r}
run2 <- run_pipeline(sheet_csv, cfg)
run2$samples$sim1$align$status
```

## Individual steps

Every step is also an exported function usable on its own, always with an
`extra_args` escape hatch to pass any tool option:

```{r}
ref <- file.path(tempdir(), "genome.fa")
invisible(file.copy(system.file("extdata/sim/genome.fa",
                                package = "nanoflow"), ref, overwrite = TRUE))
aln <- align_minimap2(
  system.file("extdata/sim/ont_reads.fastq.gz", package = "nanoflow"),
  ref, file.path(tempdir(), "demo_align"), sample = "sim1", threads = 2,
  extra_args = c("-k", "17"))
aln
```

## Phasing against the truth

With a VCF and a BAM, WhatsHap phases variants, haplotags the BAM, and —
because the fixture has a phased truth VCF — computes switch errors:

```{r, eval = can_run && nzchar(Sys.which("whatshap"))}
unphased <- file.path(tempdir(), "unphased.vcf")
truth <- system.file("extdata/sim/truth_small.vcf", package = "nanoflow")
lines <- readLines(truth)
body <- !startsWith(lines, "#")
lines[body] <- sub("(\\d)\\|(\\d)$", "\\1/\\2", lines[body])
writeLines(lines, unphased)

ph <- phase_whatshap(unphased, aln$outputs$bam, ref,
                     file.path(tempdir(), "demo_phase"), sample = "sim1",
                     truth_vcf = truth)
read.delim(ph$outputs$compare_tsv)[, c("all_switches", "all_switch_rate")]
```

## Provenance and reporting

A run writes `provenance.json` (every command line, tool version, runtime)
into the output directory, and `render_report(run)` builds a single HTML
report across samples. On a real project you would also enable
`steps$annotate` (SnpEff/AnnotSV) and, with POD5 input and a GPU,
basecalling — see `vignette("giab-hg002")` for a real-data walkthrough,
and `?download_references` for fetching GRCh38 and databases.
